---
title: "2026-10-05_p8157_hw-01"
author: "Chris"
date: "2026-10-05"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    theme: cosmo
    css: style.css
---

The full homework can be found here: https://github.com/ChrisW12372/2026-10-05_p8157_hw-01.git

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(plotly)
library(knitr)

load("data/Six Cities.RData")
```


# Question b

## EDA for topeka

### overall checking

```{r}
df_topeka <- topeka |>
  arrange(id, age) |>
  mutate(
    height = as.numeric(height),
    age = as.numeric(age),
    height.init = as.numeric(height.init),
    age.init = as.numeric(age.init)
  )
```


```{r}
df_topeka_summary <- df_topeka |>
  group_by(id) |>
  summarise(
    n_visits = n(),
    age_first = first(age),
    age_last = last(age),
    follow_up_time = age_last - age_first,
    height_fitst = first(height),
    log_fev1_first = first(log.FEV1),
    .groups = "drop"
  )

write_csv(df_topeka_summary, "results/2026-10-05_topeka-summary.csv")
```


```{r}
# check for NA
colSums(is.na(df_topeka))

# check for duplicated values within one age
df_topeka |>
  count(id, age, name = "n_records") |>
  filter(n_records > 1)

# non-positive follow-up time
df_one_visit <- 
  df_topeka_summary |>
  filter(follow_up_time <= 0)

write_csv(df_one_visit, "results/2026-10-05_topeka_subjects-one-visit.csv")

# whether the first measurements of age and height match the initial values 
df_baseline_check <- 
  df_topeka |>
  group_by(id) |>
  summarise(
    age_init_match = all(abs(first(age) - age.init) == 0),
    height_init_match = all(abs(first(height) - height.init) == 0),
    .groups = "drop"
  )

write_csv(df_baseline_check, "results/2026-10-05_topeka_baseline-match.csv")
```



```{r}
describe_row <- function(label, x) {
  x <- x[!is.na(x)]

  tibble(
    Characteristic = label,
    N = length(x),
    `Mean (SD)` = sprintf(
      "%.2f (%.2f)", mean(x), sd(x)
    ),
    `Median [Q1, Q3]` = sprintf(
      "%.2f [%.2f, %.2f]",
      median(x),
      quantile(x, 0.25),
      quantile(x, 0.75)
    ),
    `Min–Max` = sprintf(
      "%.2f–%.2f", min(x), max(x)
    )
  )
}
```


```{r}
overview_table <- bind_rows(
  describe_row(
    "Measurements per girl",
    df_topeka_summary$n_visits
  ),
  describe_row(
    "Age at first observation (years)",
    df_topeka_summary$age_first
  ),
  describe_row(
    "Age at last observation (years)",
    df_topeka_summary$age_last
  ),
  describe_row(
    "Observed follow-up span (years)",
    df_topeka_summary$follow_up_time
  ),
  describe_row(
    "Height at first observation",
    df_topeka_summary$height_fitst
  ),
  describe_row(
    "Log(FEV1) at first observation",
    df_topeka_summary$log_fev1_first
  ),
  describe_row(
    "Age across all observations (years)",
    df_topeka$age
  ),
  describe_row(
    "Height across all observations",
    df_topeka$height
  ),
  describe_row(
    "Log(FEV1) across all observations",
    df_topeka$log.FEV1
  )
)

kable(
  overview_table,
  caption = "Table 1. Data and follow-up characteristics of the Topeka sample",
  col.names = c(
    "Characteristic",
    "N",
    "Mean (SD)",
    "Median [Q1, Q3]",
    "Min–Max"
  ),
  align = c("l", "r", "r", "r", "r"),
  row.names = FALSE
)
```


The data include 1,994 measurements from 300 girls. Each girl has 1–12 measurements, and 48 girls (16%) have only one. I will keep all girls in the analysis and use a model that can handle different numbers of measurements. Girls with only one measurement can help estimate average lung function, but cannot show how it changes within a person.  

Girls were first observed at different ages, from 6.43 to 14.07 years. I will use actual age rather than visit number to study lung growth. I will also check how many observations are available at each age and whether girls with shorter follow-up differ from those with longer follow-up.  

Initial age and height match the first recorded values for all girls. The lowest log(FEV1) is about −0.69. I will check this value against the girl’s age and height before deciding whether it is unusual.  


### make some plots

#### age ~ log(FEV1)

```{r}
fig1_topeka_spaghetti <- 
  df_topeka |>
  ggplot(aes(x = age, y = log.FEV1)) +
  geom_line(aes(group = id), color = "#32BFAD", alpha = 0.25, linewidth = 0.4) +
  geom_point(color = "#3810A3", alpha = 0.25, size = 0.75) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "#0072B2") +
  labs(
    title = "Lung function trajectories",
    x = "Age (years)",
    y = "Log(FEV1)"
  ) + theme_minimal()

ggsave("plots/2026-10-05_fig01_topeka_spaghetti-plot.png", fig1_topeka_spaghetti,
       height = 6, width = 9, dpi = 1000)

fig1_topeka_spaghetti
```


```{r}
# find out the outlier
outlier_topeka <- df_topeka |>
  slice_min(log.FEV1, n = 1, with_ties = TRUE)

outlier_topeka
```

```{r}
# checking all measurements of id 197
df_topeka |>
  filter(id == 197) |>
  arrange(id, age)
```


Log(FEV1) generally increases with age and levels off around ages 15–16. I will therefore consider a nonlinear age effect in later models. Lung function also differs across girls, so I will explore within-person correlation.  

Subject 197 has a clearly low value and only one measurement, so I cannot compare it with her other visits. I will retain this observation for now, and later decide whether to include it into the analysis.  

#### height ~ log(FEV1)

```{r}
fig2_topeka_points <- 
  df_topeka |>
  ggplot(aes(x = height, y = log.FEV1)) +
  geom_point(color = "#DB662C", alpha = 0.35, size = 0.75) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, colour = "#8A2278") +
  geom_point(data = outlier_topeka, colour = "#5a6acd") +
  geom_text(data = outlier_topeka, colour = "#E6DD39", aes(label = paste0("ID", id)), vjust = -1) +
  labs(
    title = "Lung function and height",
    x = "Height",
    y = "Log(FEV1)"
  ) + theme_minimal()

ggsave("plots/2026-10-05_fig02_topeka_point-plot.png", fig2_topeka_points,
       height = 6, width = 9, dpi = 1200)

fig2_topeka_points
```


#### age ~ height


```{r}
fig3_topeka_age_height_spaghetti <- 
  df_topeka |>
  ggplot(aes(x = age, y = height)) +
  geom_line(aes(group = id), color = "#7FBA38", alpha = 0.25, linewidth = 0.4) +
  geom_point(color = "#6EAFDB", alpha = 0.35) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "#894CC7") +
  labs(
    title = "Height trajectories",
    x = "Age (years)",
    y = "Height"
  ) + theme_minimal()

ggsave("plots/2026-10-05_fig03_topeka_age-height-spaghetti.png", fig3_topeka_age_height_spaghetti, height = 6, width = 9, dpi = 1000)

fig3_topeka_age_height_spaghetti
```


Age has a strong relationship with height, and therefore, the increase of log(FEV1) by age may have something to do with height.  






